---
title: "Data Visualizatio in R"
author: "Alin Morariu"
date: "20/09/2021"
output: html_document
---
## Section 0
**Disclaimer:** this section as a pre-introduction and serves in introduce the data types and various objects we will be working with in R. 


R is open sourced software/computing language. It has a simple syntax which is easy to read and write making it popular amongst scientists and STEM professionals. There are several types of documents we can work in to take advantage of R. The first is called an R script and has a .R file type. This is a blank document in which you write R code and then execute. The second is an Rmarkdown document which have a .Rmd file type. Rmarkdown documents allow you to both write R code and word process in markdown (think of it as simpified Word format). In these documents, the output of your code is shown alongside text making them great options for writing analysis reports to avoid saving jpeg's and importing them to a Word document. 

### R Objects 

```{r R objects }
# storing variables (use the # sign to comment your code, this is code that does NOT get run)
x0 <- 42
class(x0) # check the object type 
```
R has 2 ways to assign values to a variable in memory, namely the arrow ( <- ) and the equal sign ( = ). A good convention is to use the arrow ( <- ) to assign new variables while using the equal sign ( = ) to set parameter values. 
```{r}
# can store multiple values in the same memory slot 
v0 <- c(41, x0, 43)
v0

# acccessing specific parts of the vector 
v0[2]
v0[1:3]

# can store multiple vectors 
df0 <- data.frame(v0, 2*v0, v0/2)
df0
```


### R operations 

```{r R operations}
# add and substract
x0 + x0 
x0-3

# mult + div 
x0*4
x0/4

v0*4

# mult with dataframes 
2*df0
```

```{r Matrix Operations}
# define a 3x3 matrix 
mt0 <- as.matrix(df0)
mt0

t(mt0)
x0^2
# vector multiplication 
v0 * v0 # element wise

v0 %*% v0 # dot product 

v0 %*% t(v0) # vector product 

mt0 %*% mt0 # matrix product 
```

### Functions

Sometimes we want to do a defined set of operations and this is where user functions come in. R contains a set of built in functions which come in handy, but we can also define our own functions. Writing your own functions can help you save a lot of time by automating any calculations or step you have to redo often. 

```{r Function}
sum(1,2,3)
sum(v0)

my_sum <- function(vector){
     # sum the first and last entry of a vector 
     x <- vector[1] + vector[length(vector)]
     
     return(x)
}

my_sum(v0)
my_sum(c())
```

## Section 1
Since R is open source, it has many people who write code and make it available to the public. Developers wrap that code up as a package so you and I can download and use their useful functions. Before using a pacakge you must first install it with the `install.packages(pkg_name)` command. This only needs to be done *once* on your computer. The package must then be loaded with the `library('pkg_name)` command. This needs to be done for every new file. 
```{r}
# install.packages(tidyverse) - ONLY INSTALL PACKAGES ONCE
library(tidyverse)  # LOAD PACKAGES FOR EVERY FILE - to load a package call it with the library command
library(lubridate)
library(scales)
```

### Loading data 
```{r Previewing data}
time_series_covid19_confirmed_global <- read_csv("time_series_covid19_confirmed_global.csv")

head(time_series_covid19_confirmed_global) # show the top rows
tail(time_series_covid19_confirmed_global) # show the bottom rows
```

### Reshaping Data
Turn our data into a tidy format (each variable is its own column, each observation is its own row). 
```{r}
infected_data_tidy <- time_series_covid19_confirmed_global %>% 
     gather(date_stamp, # this is new column which will take values from the original dataset columns, 
            num_infect, # a new value representing the number of confirmed/infected covid cases
            '1/22/20':'9/19/21') %>% 
     select(Country = 'Country/Region', Province = 'Province/State', date = 'date_stamp', num_infect = 
                 num_infect)

infected_data_tidy$date <- lubridate::mdy(infected_data_tidy$date)
tail(infected_data_tidy, 10)
```
Now that we have data that is clean, we can move on to create some useful/interesting summaries. 

```{r Summarize data}
# finding single number summaries like min, max, mean, sd, etc
infected_data_tidy %>% summarise(max(num_infect))
```

### Filtering and Selecting 
Filtering data refers to finding row entries which satisfy a specific (logical) condition. Selecting data refers to picking out columns (or variables) within a dataset.
```{r}
infected_data_tidy %>% filter(num_infect == 42087432)
# or 
infected_data_tidy %>% filter(Country == 'Canada', Province == 'Ontario')
```

### Summary Tables
Summary tables are custom dataframes containing specific information that is calculated based on the data. They can be very simple such as the one below but can get much more complicated based on the specificity of the summary required. 
```{r}
infected_data_tidy %>% summarise(Min = min(num_infect),
                                 Max = max(num_infect),
                                 Avg = mean(num_infect),
                                 Median = median(num_infect), 
                                 Std_dev = sd(num_infect))
```
We can also go deeper and be more granular. 
```{r}
infected_data_tidy %>% 
     group_by(Country) %>% 
     summarise(Min = min(num_infect),
                                 Max = max(num_infect),
                                 Avg = mean(num_infect),
                                 Median = median(num_infect), 
                                 Std_dev = sd(num_infect)) %>% 
     filter(Country %in% c('Canada', 'Iran', 'United Kingdom', 'US')) %>% 
     arrange(desc(Avg))
```

### Grouped Summaries
One of the most powerful tools within dplyr is the `group_by` function because it allows you to do calculations on subsections within the data. In our case, we can group our data by the country, province, or date and summarize it accordingly. 
```{r}
infected_data_tidy %>% 
     group_by(date) %>% 
     summarise(Min = min(num_infect),
               Max = max(num_infect),
               Avg = round(mean(num_infect),2),
               Infect = sum(num_infect),
               Median = median(num_infect), 
               Std_dev = sd(num_infect))
```

## Section 2 

### Combining Datasets 
Another useful action is combining datasets. So far we have only seen the number of infected but we also have datasets for the number of deaths and recoveries. Let's load those in as well. 

```{r}
death_data_tidy <- read_csv("time_series_covid19_deaths_global.csv")  %>% 
     gather(date_stamp, # this is new column which will take values from the original dataset columns, 
            num_dead, # a new value representing the number of confirmed/infected covid cases
            '1/22/20':'9/19/21') %>% 
     select(Country = 'Country/Region', Province = 'Province/State', date = 'date_stamp', num_dead = 
                 num_dead)

tail(death_data_tidy)
```
Now we have 3 data sets that have 3 different numbers of interest. However, note that they all share 2 columns: Country and date. We will use those two columns to match up entries and get a complete dataframe. Think of this like a vlookup in excel

```{r}
# first we need to get the new datasets into the same format 
death_data_tidy$date <- lubridate::mdy(death_data_tidy$date)

# now we join them! 
complete_data <- full_join(infected_data_tidy, death_data_tidy) %>% 
     group_by(Country) %>% 
     mutate(num_recovered = num_infect - num_dead, 
            days = date - first(date) + 1)
tail(complete_data)
```

Often times we need to construct custom data frames to generate plots we are interested in. Say we want to map the daily case counts for Canada as a country. 
```{r}
complete_data %>% 
     filter(Country %in% c('Canada')) %>%
     group_by(date) %>% 
     summarise(Country = last(Country),
               cum_infect = sum(num_infect),
               cum_dead = sum(num_dead), 
               cum_recovered = sum(num_recovered),
               days = last(days)) %>%
     tail(8)

#      ggplot(aes(x = date)) + 
#           geom_line(aes( y = c(0,diff(cum_infect)))) + 
#           geom_smooth(aes( y = c(0,diff(cum_infect))), span = 0.1, se= FALSE) + 
#           geom_line(aes( y = cum_infect/days), colour = 'gold') + 
#           geom_vline(xintercept = as_date('2021-07-17'), linetype = 4, colour = 'red', show.legend =  TRUE) + 
#           geom_text(aes(x = as_date('2021-07-17'), label = 'day I graduated', y = 9000), colour = 'red', angle = 90, vjust = -1, text=element_text(size=11)) + 
#      scale_y_continuous(labels = comma) 
```

## Section 3
### Making plots 
ggplot works by layering different parts of a plot. Starting with the data, you lay down the X and Y axis (potentially Z), and keep adding elements to finish the plot. This first example will be done very slowly, then we will combine several lines to make it faster. 

```{r}
# take the previous data set and plot it 
complete_data %>% 
     filter(Country %in% c('Canada')) %>%
     group_by(date) %>% 
     summarise(Country = last(Country),
               cum_infect = sum(num_infect),
               cum_dead = sum(num_dead), 
               cum_recovered = sum(num_recovered),
               days = last(days)) %>%
     ggplot() + 
     aes(x = date) + 
     aes(y = cum_infect) +
     geom_line() +
     labs(x = 'Date') + 
     labs(y = 'Number of Cases') + 
     labs(title = 'Covid-19 Cases in Canada') + 
     labs(subtitle = 'Confirmed Covid-19 cases in Canada based on Health Canada counts')+ 
     scale_y_continuous(labels = comma) +
     theme_minimal()

# or more concisely
complete_data %>% 
     filter(Country %in% c('Canada')) %>%
     group_by(date) %>% 
     summarise(Country = last(Country),
               cum_infect = sum(num_infect),
               cum_dead = sum(num_dead), 
               cum_recovered = sum(num_recovered),
               days = last(days)) %>%
     ggplot(aes(x= date, y = cum_infect)) + 
     geom_line(linetype = 'longdash', colour = 'green') + 
     labs(x = 'Date', y = 'Number of Cases', 
          title = 'Cumulative Covid-19 Cases in Canada', 
          subtitle = 'Confirmed Covid-19 cases in Canada based on Health Canada counts') + 
     scale_y_continuous(labels = comma) +
     theme_dark()
```

```{r}
# what about daily counts?
```


```{r}
complete_data %>%
     filter(Country %in% c('Canada')) %>%
     group_by(date) %>%
     summarise(
          Country = last(Country),
          cum_infect = sum(num_infect),
          cum_dead = sum(num_dead),
          cum_recovered = sum(num_recovered),
          days = last(days)
     ) %>%
     ggplot(aes(x = date)) +
     # lines on the graph
     geom_line(aes(y = c(0, diff(cum_infect))), linetype = 'solid', colour = 'black') +
     geom_smooth(aes(y = c(0, diff(cum_infect))),
                 span = 0.2,
                 se = FALSE,
                 linetype = "longdash") +
     geom_vline(
          xintercept = as_date('2021-07-17'),
          linetype = 4,
          colour = 'red',
          show.legend =  TRUE
     ) +
     geom_text(
          aes(
               x = as_date('2021-07-17'),
               label = 'day I graduated',
               y = 9000
          ),
          colour = 'red',
          angle = 90,
          vjust = -1,
          text = element_text(size = 11)
     ) +
     # formatting
     labs(
          x = 'Date',
          y = 'Number of Cases',
          title = 'Covid-19 Cases in Canada',
          subtitle = 'Confirmed Covid-19 cases per day in Canada based on Health Canada counts'
     ) +
     scale_y_continuous(labels = comma) +
     theme_minimal()
```
What if we want to break this up by province? 
```{r}
complete_data %>%
     filter(Country %in% c('Canada'), 
            Province %in% c("Alberta", 'British Columbia', "Ontario", 'Quebec')
            ) %>% 
     ggplot() + 
     aes(x = date) +
     geom_line(aes( y = num_infect, colour = 'Infected')) +
     geom_line(aes(y = num_dead, colour = 'Dead')) + 
     scale_color_manual(name = 'Legend', 
                        values = c('Infected' = 'red',
                                   'Dead' = 'black')) + 
     facet_wrap(~ Province) + 
     geom_vline(
          xintercept = as_date('2021-07-17'),
          linetype = 4,
          colour = 'darkgreen',
          show.legend =  TRUE
     ) +
     labs(
          x = 'Date',
          y = 'Number of Cases',
          title = 'Covid-19 Cases in Canada',
          subtitle = 'Confirmed Covid-19 cases per day in Canada based on Health Canada counts'
     ) +
     scale_y_continuous(labels = comma) +
     guides() + 
     theme_light()
     
```

Some other plots
```{r Bar Graph}
complete_data %>%
     filter(Country %in% c('Canada', 'Iran', 'United Kingdom', 'Venezuela', 'Zimbabwe')) %>% 
     summarise(total_cases = max(num_infect)) %>% 
     ggplot(aes(x = fct_reorder(Country, total_cases), y = total_cases)) + 
     geom_col(fill = 'steelblue', colour = 'navyblue', alpha = 0.7) +
     labs(
          x = 'Country',
          y = 'Number of Cases',
          title = 'Total Covid-19 Cases Counts',
          subtitle = 'Confirmed Total Covid-19 cases in several countries'
     ) +
     scale_y_continuous(labels = comma) +
     coord_flip()+
     theme_light()
     
```

```{r Boxplots }
complete_data %>%
     filter(Country %in% c('Canada', 'Iran', 'United Kingdom', 'Argentina', 'South Africa')) %>%
     group_by(date, Country) %>%
     summarize(
          total_infect = sum(num_infect),
          total_dead = sum(num_dead),
          days = last(days)
     ) %>%
     ggplot() +
     aes(x = Country) +
     aes(y = total_infect) +
     geom_jitter(alpha = .4,
                 height = 0,
                 width = .25,
                 show.legend = FALSE) +
     aes(col = Country) +
     geom_boxplot(alpha = 0.25, show.legend = FALSE) +
     aes(fill = Country) +
     labs(
          x = 'Country',
          y = 'Number of Cases',
          title = 'Boxplots for Covid-19 Cases Counts',
          subtitle = 'Confirmed Total Covid-19 cases in several countries'
     ) +
     scale_y_continuous(labels = comma) +
     coord_flip() +
     theme_light()
```

