---
title: "R_workshop_Feb_2021"
author: "Alin Morariu"
date: "25/02/2021"
output: html_document
---

# Welcome! 

This is a 2 part workshop focusing on key coding skills in R that will be useful in a research setting. Topics covered will include 

* R objects
* Basic operations
* Introduction to knitr
* Introduction to the tidyverse
* Data wrangling (via dplyr)
* Data visualization (via ggplot2)
* Export our results to a PDF via an RMarkdown file 

### What is R? 

R is a statistical program language that gives you a means of computing key values and plots to analyze your data. We will be using an IDE overlayed onto R called R Studio which provies a clean, intuitive environment for you to work in. R Studio has 4 main panes - Source file, Console, Environment, and Files. 

### R Objects 

```{r R objects }
x0 <- 42
class(x0)
```

```{r}
# can store multiple values in the same memory slot 
v0 <- c(41, x0, 43)
v0

# acccessing specific parts of the vector 
v0[2]
v0[1:3]

# can store multiple vectors 
df0 <- data.frame(v0, 2*v0, v0/2)
df0
```


### R operations 

```{r R operations}
# add and substract
x0 + x0 
x0-3

# mult + div 
x0*4
x0/4

v0*4

# mult with dataframes 
2*df0
```

```{r Matrix Operations}
# define a 3x3 matrix 
mt0 <- as.matrix(df0)
mt0

t(mt0)
x0^2
# vector multiplication 
v0 * v0 # element wise

v0 %*% v0 # dot product 

v0 %*% t(v0) # vector product 

mt0 %*% mt0 # matrix product 
```

### Functions

Sometimes we want to do a defined set of operations and this is where user functions come in. R contains a set of built in functions which come in handy, but we can also define our own functions.

```{r Function}
sum(1,2,3)
sum(v0)

my_sum <- function(vector){
     # sum the first and last entry of a vector 
     x <- vector[1] + vector[length(vector)]
     
     return(x)
}

my_sum(v0)
my_sum(c())
```

### Packages 

```{r packages, eval = FALSE }
install.packages('tidyverse')
install.packages('knitr')
```

```{r call the packages}
library(tidyverse)
library(knitr)
library(scales)
```

Today we will focus on the usage of two key R packages - tidyverse and knitr. 

### Tidyverse

The tidyverse is a conglometration of smaller packages created with the goal of simplifying many of R's data science capabilities as well as taking advantage of modern computing power through highly efficient back end code (you do NOT need to worry about this and this is why having it in a package is so convenient). Within the tidyverse, we will focus on 2 main sub-packages. Namely, dplyr (a package used for data wrangling and cleaning), and ggplot2 (a packaged used for data visualization and creation of plots). Both of these packages have wonderful cheatsheets online which I strongly recommend using (I open these up every time I do anything in R). 

* dplyr: https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf
* ggplot2: https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf 

### Knitr

Knitr will be particularly useful to us since it will allow us to export our work from these .Rmd files into usable, and sometimes publishable, documents. The RMarkdown documents can be exported to HTML, PDF, or Word documents making them very useful when working on a project. The alternative is to write all of your code in an R script (a .R file), and work with the outputted objects individually (for example, you can save any of your plots as .jpeg or .png files which you can then slot into your figures in another document). I recommend using the RMarkdown setting due to it's ability to handle inline code via the code chunks you've been seeing so far. 

### Playing with Covid-19

Let's load in some data which we downloaded from the Johns Hopkins Covid-19 dashboard. The most recent files are available for download [here](https://github.com/CSSEGISandData/COVID-19). 

```{r}
# need to set where we get our data from -CHANGE THIS FOR YOUR COMPUTER
setwd("~/Documents/GitHub/R-Workshop-for-STEM/2021")

# many ways to import data
covid_global <- read_csv("time_series_covid19_confirmed_global.csv")
covid_US <- read_csv("time_series_covid19_confirmed_US.csv")
```

Now we need to view our data and decide on what to do with it! 

```{r Viewing our data}
head(covid_US)
```
The data contains information about the location and for each location, a time series (data indexed by a time/date) for the number of confirmed Covid-19 cases between January 1st, 2020 and November 28th, 2020. Before we do anything more, we always want to convert the data into a *tidy* format. Tidy data has a unique variable  in each column and one observation per row. Our data frame has many measurements for the number of cases across the rows and must therefore be transformed into tidy form. To do so we use the `r gather` function. It is incredibly powerful because it lets us map each column into a row with the use of key-value pairings. The idea is that we want each row to contain the country, province, city, date, *and* the number of confirmed cases. Having data in that format will allow us to group it together based on the different levels and aggregate as needed. The code is shown below
```{r}
infected_data_tidy <- covid_US %>% 
     gather(date_stamp, # this is the new column which will take values from covid_US columns
            num_infect, # this is the new value 
            '1/22/20':'11/28/20' # the columns we're getting values from
            ) %>% 
     rename(Province = `Province_State`, Country = `Country_Region`, City = `Admin2`) %>% 
     select(Country, Province, date_stamp, num_infect) %>% 
     group_by(Province)
head(infected_data_tidy)
```
We also want to change the dates to be of the data type instead of the strings they currently are. The lubridate package (part of the tidyverse package) helps us do that with the various date formats (month-day-year in this case). Since we only want to change a specific column, we can use the $ operator which subsets the dataframe (column-wise). 

```{r}
# change dates from string format to datetime format
infected_data_tidy$date_stamp <- lubridate::mdy(infected_data_tidy$date_stamp)
```

Total number of infections 
```{r}
# we want to aggregate data on a country scale for summary purposes
infected_summary <- infected_data_tidy %>%
     group_by(date_stamp) %>%                           # each locale has a date stamp and we want to summarize by the day
     summarise(total = sum(num_infect),
               total_per_thousand = sum(num_infect)/1000,
               avg_infected = mean(num_infect),
               sd_infected = sd(num_infect))                 # aggregating function

tail(infected_summary) # a variant of head
```

```{r}
# goal: find the max number of infections in a single day
max(infected_data_tidy$num_infect)

# built in functions which help sumamrize data 
summary(infected_data_tidy$num_infect)

```

Let's make some plots.

```{r}
infected_summary %>% ggplot(aes(x = date_stamp, y = total)) + 
        geom_line(colour =  'maroon') + 
        labs(title = 'US Total Number of Covid-19 Cases',
             x = 'Date',
             y = 'Total Confirmed Cases') + 
        scale_y_continuous(labels = comma)

infected_summary %>% ggplot(aes(x = date_stamp, y = sd_infected)) + 
        geom_line(colour =  'darkgreen') + 
        labs(title = 'US Total Number of Covid-19 Cases',
             x = 'Date',
             y = 'Total Confirmed Cases') + 
        scale_y_continuous(labels = comma)
```

```{r}
infected_summary %>% ggplot(aes(x = date_stamp, y = c(0,diff(total)))) + 
        geom_line(colour =  'maroon') + 
        geom_smooth(method = 'loess', se = FALSE) + 
        labs(title = 'US Daily Number of Covid-19 Cases',
             x = 'Date',
             y = 'Total Confirmed Cases') + 
        scale_y_continuous(labels = comma)
```

```{r}
# option 1: build a histogram to see distribution
infected_summary %>% ggplot(aes(c(0,diff(total)))) + 
        geom_histogram(bins = 25, fill = 'steelblue', colour = 'navy') + 
        labs(title = 'Daily Infection Totals',
             subtitle = 'Within the USA',
             x = 'Number of Cases',
             y = 'Number of Days') +
        scale_x_continuous(labels = comma) +
        theme_minimal() # theme depends on your preference

# option 2: kernel density 
infected_summary %>% ggplot(aes(c(0,diff(total)))) + 
        geom_density(kernel = 'gaussian', fill = 'steelblue', colour = 'navy') + 
        labs(title = 'Daily Infection Totals',
             subtitle = 'Within the USA',
             x = 'Number of Cases',
             y = 'Number of Days') +
        scale_x_continuous(labels = comma) +
        scale_y_continuous(labels = comma) +
        theme_minimal() # theme depends on your preference
```

```{r}
infected_data_tidy %>% 
        group_by(Province) %>% 
        summarise(total_infected_per_state = sum(num_infect)) %>%
        top_n(10) %>% 
        ggplot(aes(x = reorder(Province, -total_infected_per_state), y = total_infected_per_state)) +
                geom_bar(stat = 'identity', fill = 'steelblue', colour = 'navy') + 
        labs(title = 'Total number of Covid-19 cases',
             subtitle = 'Top 10 MOST affected states',
             x = 'State',
             y = 'Number of Confirmed Cases') +
        scale_y_continuous(labels = comma) +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45))

infected_data_tidy %>% 
        group_by(Province) %>% 
        summarise(total_infected_per_state = sum(num_infect)) %>%
        top_n(-10) %>% 
        ggplot(aes(x = reorder(Province, -total_infected_per_state), y = total_infected_per_state)) +
                geom_bar(stat = 'identity', fill = 'steelblue', colour = 'navy') + 
        labs(title = 'Total number of Covid-19 cases',
             subtitle = 'Top 10 LEAST affected states',
             x = 'State',
             y = 'Number of Confirmed Cases') +
        scale_y_continuous(labels = comma) +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45))
```

```{r}
infected_data_tidy %>%
        ggplot(aes(date_stamp, num_infect)) + 
        geom_line(colour = 'maroon') + 
        facet_wrap(~Province) + 
        scale_y_continuous(labels = comma) + 
        labs(title = "US Total case per State for Covid-19",
             x = "Date",
             y = "Totals") +
        theme_light()
```

